A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-09-29 latex/main.tex
Move comparison with other methods to supplement
Closes #654

By automatically detecting and incorporating recombination events
into the ARG, sc2ts provides us with a systematic means of 
discovering and assessing evidence around these events.
Table~\ref{tab:pango_x_lineages} summarises
these results for the 44 Pango X lineages present in
the ARG (XAC and XAD have been omitted, and XM 
partially represented, corresponding to 23 samples in total; 
see Document~\ref{sec:supplement_x_pango_lineages}).
For 16 Pango X lineages the oldest node that has been assigned
the focal Pango lineage is a recombination node, and that node 
is the ancestor of all samples assigned that (or a descendant) Pango 
lineage. These are the Type I events, in which the ARG reflects the
expected phylogenetic structure of Pango designation.
Among these events, concordance with the community-consensus 
parent lineages and breakpoint intervals is excellent (see next 
section). 13 of the Type I events are monophyletic for the focal
pango (i.e., the descendants are entirely from that lineage); 
Figure~\ref{fig:subgraphs}A shows XA as an example (see also 
Document~\ref{sec:jackson_recombs} for further analysis 
of the Jackson et al.\ recombinants).

@@

\subsection*{Comparison with other recombination detection methods}

%%% TODO: Mention RIPPLES somewhere here? Or just in Discussion?
We compared sc2ts against three state-of-the-art methods that detect inter-lineage recombinants:
RecombinHunt \cite{Alfonsi2024}, CovRecomb \cite{Li2024CovRecomb}, and rebar (https://github.com/phac-nml/rebar).
These methods use the same underlying approach:
they scan candidate recombinant sequences for mutations 
associated with different Pango lineages
without using a phylogeny.

\begin{table}
\caption{
Concordance among methods in characterizing Pango X lineages.
These comparisons focus on the Pango X lineages associated with
the Type 1 recombination events (Table~\ref{tab:pango_x_lineages})
For each method, we calculated the number of Pango X lineages
which have inferred parent Pango lineages and breakpoint intervals
concordant with those proposed by the community.
The denominators indicate the number of Pango X lineages
for which detection results are available from the original studies \cite{Alfonsi2024,Li2024CovRecomb}.
For the Pango X lineages with non-overlapping breakpoint intervals,
the distances between the inferred breakpoint intervals and 
those proposed by the community are shown (median and range).
}
\label{tab:method_concordance}
\begin{tabular}{lccc}
\toprule & \multicolumn{1}{c}{Parent lineages} & \multicolumn{2}{c}{Breakpoint intervals} \\
\cmidrule(lr){2-2} \cmidrule(lr){3-4}
Method & Concordant (\%) & Concordant (\%) & Distance (bases) \\
\midrule
Sc2ts & 14 / 16 (87.5) & 14 / 16 (87.5) & 540.5 (315, 766)\\
RecombinHunt-GISAID & 13 / 15 (86.7) & 10 / 15 (66.7) & 764.0 (4, 1202)\\
RecombinHunt-Nextstrain & 12 / 15 (80.0) & 11 / 15 (73.3) & 1160.0 (559, 2785)\\
CovRecomb & 7 / 8 (87.5) & 4 / 8 (50.0) & 1562.5 (217, 13208)\\
\bottomrule
\end{tabular}
\end{table}

@@

\subsubsection*{Comparison with recombination detection methods}

@@

>>>>

By automatically detecting and incorporating recombination events
into the ARG, sc2ts provides us with a systematic means of 
discovering and assessing evidence around these events.
Table~\ref{tab:pango_x_lineages} summarises
these results for the 44 Pango X lineages present in
the ARG (XAC and XAD have been omitted, and XM 
partially represented, corresponding to 23 samples in total; 
see Document~\ref{sec:supplement_x_pango_lineages}).
For 16 Pango X lineages the oldest node that has been assigned
the focal Pango lineage is a recombination node, and that node 
is the ancestor of all samples assigned that (or a descendant) Pango 
lineage. These are the Type I events, in which the ARG reflects the
expected phylogenetic structure of Pango designation.
Among these events, concordance with the community-consensus 
parent lineages and breakpoint intervals is excellent,
with similar performance to
the RecombinHunt \cite{Alfonsi2024}
and CovRecomb \cite{Li2024CovRecomb} detection methods
(Table~\ref{tab:method_concordance}).
Of the Type I events, 13 are monophyletic for the focal
pango (i.e., the descendants are entirely from that lineage); 
Figure~\ref{fig:subgraphs}A shows XA as an example (see also 
Document~\ref{sec:jackson_recombs} for further analysis 
of the Jackson et al.\ recombinants).

@@

@@

\subsubsection*{Comparison with recombination detection methods}
[TODO Merge in this text that was in the Results]

@@

\begin{table}
\caption{
Concordance among methods in characterizing Pango X lineages.
These comparisons focus on the Pango X lineages associated with
the Type 1 recombination events (Table~\ref{tab:pango_x_lineages})
For each method, we calculated the number of Pango X lineages
which have inferred parent Pango lineages and breakpoint intervals
concordant with those proposed by the community.
The denominators indicate the number of Pango X lineages
for which detection results are available from 
the original studies \cite{Alfonsi2024,Li2024CovRecomb}.
For the Pango X lineages with non-overlapping breakpoint intervals,
the distances between the inferred breakpoint intervals and 
those proposed by the community are shown (median and range).
}
\label{tab:method_concordance}
\centering
\begin{tabular}{lccl}
\toprule & \multicolumn{1}{c}{Parent lineages} & \multicolumn{2}{c}{Breakpoint intervals} \\
\cmidrule(lr){2-2} \cmidrule(lr){3-4}
Method & Concordant (\%) & Concordant (\%) & Distance (bases) \\
\midrule
Sc2ts & 14 / 16 (87.5) & 14 / 16 (87.5) & 540.5 (315, 766)\\
RecombinHunt-GISAID & 13 / 15 (86.7) & 10 / 15 (66.7) & 764.0 (4, 1202)\\
RecombinHunt-Nextstrain & 12 / 15 (80.0) & 11 / 15 (73.3) & 1160.0 (559, 2785)\\
CovRecomb & 7 / 8 (87.5) & 4 / 8 (50.0) & 1562.5 (217, 13208)\\
\bottomrule
\end{tabular}
\end{table}
